---
title: "Thinking longitudinally"
subtitle: "Intro to data and motivation"
format:
  revealjs: 
    scrollable: true  
    slide-number: true
    chalkboard: 
      buttons: false
    preview-links: auto
    css: styles.css
    footer: <https://alda22.netlify.app>
resources:
  - demo.pdf
---

## Goals for today:

```{r, echo=FALSE}
library(tidyverse)
```

-   Get a feeling for how to think/talk about longitudinal/repeated measures data

-   Introduce some important terms

-   Begin to develop a framework for analysis

## how to think longitudinal-y

::: columns
::: {.column width="30%"}
1.  lines/ trajectories

2.  variance decomposition
:::

::: {.column width="70%"}
```{r, echo = FALSE}
#| fig-width: 10
#| fig-height: 7.5

simp<- tribble(
  ~ID,  ~Y, ~time,
1,5,1,
1,7,2,
2,4,1,
2,6,2,
3,3,1,
3,5,2,
4,2,1,
4,4,2,
5,1,1,
5,3,2)

ggplot(simp, aes(x=time, y=Y)) +
    geom_point(size=5) + theme(text = element_text(size = 20)) 
```
:::
:::

## How do we define "change"?

-   Types of change (most common):

    -   Differential / rank order consistency/rank order stability (correlations)
    -   Mean level/ absolute change (mean differences)

------------------------------------------------------------------------

::: columns
::: {.column width="50%"}
Perfect rank order, mean level increase
:::

::: {.column width="50%"}
No rank order, mean level increase
:::
:::

```{r,message = FALSE, warning = FALSE}
library(tidyverse)

 ro.ml <- ggplot(simp, aes(x=time, y=Y)) +
    geom_point(size=5) + 
  stat_summary(fun = mean, geom="line", size = 4)+ geom_smooth(aes(group = ID), method=lm, se=FALSE) + theme(text = element_text(size = 20))

simp2<- tribble(
  ~ID,  ~Y, ~time,
1,1,1,
1,5,2,
2,1.5,1,
2,4.5,2,
3,2,1,
3,4,2,
4,2.5,1,
4,3.5,2,
5,3,1,
5,3,2)

noro.ml<- ggplot(simp2, aes(x=time, y=Y)) +
    geom_point(size=5) +   stat_summary(fun = mean, geom="line", size = 4) + geom_smooth(aes(group = ID), method=lm, se=FALSE) + theme(text = element_text(size = 20))

library(patchwork)
ro.ml + noro.ml

```

## How do we defining "change"?

-   Because there are many types of change, we will view change in terms of the model.

-   (Usually) it is clearer to refer to the type of change in terms of an equation or pictorially. Putting a word onto it usually causes confusion, which is why there are a lot of redundant terms in the literature.

-   Models may be able to tell us about two different types of change (within person vs between person change)

## Prerequisites

-   Ordinal or greater scale of measurement us easiest. Dichotomous is hard.

-   Construct has the same meaning across measurement occasions. Usually the same items. Called measurement invariance. Complicates developmental work.

-   2 or more measurement occasions. More is better! Though often 3 - 10 is practically fine for some models. With 30+ occasions you have "intensive" longitudinal data which presents new models and opportunities.

## Defining a time metric

-   Time is the most important part of a longitudinal analyses. The key to interpreting your output is to know how you handled your time variable.

-   What is the process that is changing someone? Age? Time in study? Year? Wave?

-   Is it a naturally occurring developmental process? Then maybe age is the best metric. What about tracking child's cognitive ability, something that might be influenced by level of schooling? Here grade may be more important than age. If you are running an intervention you may want to put everyone on the same starting metric and then control for nuisance variables like age or schooling level.

## Example

::: panel-tabset
### Plot

```{r, eval = FALSE}
ggplot(example,
   aes(x = year, y = SMN7, group = ID)) + geom_point()  

```

### Data

```{r, eval = FALSE, message=FALSE, warning = FALSE}

example <- read_csv("~/Box/5165 Applied Longitudinal Data Analysis/Longitudinal/example.csv")
example$year <- example$week
```
:::
